
---
title: "MiFish commands for processing Black Guillemot samples from Shoals"
author: "Gemma Clucas"
date: "2026-04-01"
output: github_document
---

## 1. Set up

```{r setup}
options(scipen=999)
knitr::opts_chunk$set(echo = TRUE, digits = 1)
knitr::opts_chunk$set(save_output = FALSE) # prevent outputs being over-written on knitting
library(tidyverse)
library(knitr)
library(qiime2R)
```

All analyses performed in the folder `Fecal_metabarcoding/Black-Guillemots`.
```
conda activate qiime2-amplicon-2024.10 
mkdir MiFish
cd MiFish
```

## 2. Import the data into QIIME2

```
for K in 24 64 81 146; do
  qiime tools import\
    --type 'SampleData[PairedEndSequencesWithQuality]'\
    --input-path /Volumes/Data_SS1/MiFish/BLGU_Shoals/Plate$K/reads/ \
    --input-format CasavaOneEightSingleLanePerSampleDirFmt\
    --output-path demux_Plate$K.qza
done
  

for K in 24 64 81 146; do
  qiime demux summarize \
    --i-data demux_Plate$K.qza \
    --o-visualization demux_Plate$K.qzv
done

``` 

## 3. Trim primers using cutadapt

The MiFish sequences are:   

F primer: GTCGGTAAAACTCGTGCCAGC (21 bp)   
R primer: CATAGTGGGGTATCTAATCCCAGTTTG (27 bp)   

### Trim 3' ends first
At the 3' end of the read, the primer will have been read through after reading the MiFish amplicon. Look for the reverse complement of the reverse primer in read 1 (—p-adapter-f) and the reverse complement of the forward primer in R2 (—p-adapter-r).

F primer reverse complement: GCTGGCACGAGTTTTACCGAC   
R primer reverse complement: CAAACTGGGATTAGATACCCCACTATG    

```
for K in 24 64 81 146; do
  qiime cutadapt trim-paired \
    --i-demultiplexed-sequences demux_plate$K.qza \
    --p-adapter-f CAAACTGGGATTAGATACCCCACTATG \
    --p-adapter-r GCTGGCACGAGTTTTACCGAC \
    --o-trimmed-sequences trimd_Plate$K.qza \
    --verbose > cutadapt_out_Plate$K.txt
done
```
To see how much data passed the filter for each sample:
```
for K in 24 64 81 146; do
  grep "Total written (filtered):" cutadapt_out_Plate$K.txt 
done
```
It looks like 77% is being kept for the first samples, but the samples on plate 146 are at 100%, suggesting a problem with them.

**Did Alex use MiniFish primers for plate 140??**

Try trimming for minifish instead

The MiniFish sequences are:   

MiniFish_Ts_F: GTTATACGAGAGGCCCAAGTTG
MiniFish_Ts_R: TAAAGCCACTTTCGTGGTTG

```
for K in 146; do
  qiime cutadapt trim-paired \
    --i-demultiplexed-sequences demux_plate$K.qza \
    --p-adapter-f CAACCACGAAAGTGGCTTTA \
    --p-adapter-r CAACTTGGGCCTCTCGTATAAC \
    --o-trimmed-sequences trimd_MiniFish_Plate$K.qza \
    --verbose > cutadapt_MiniFish_out_Plate$K.txt
done

for K in 146; do
  grep "Total written (filtered):" cutadapt_MiniFish_out_Plate$K.txt 
done
```
**Yep the MiniFish primer sequences are there.**   
Most samples have about 50-60% data passing. Going to move all of the files for Plate 146 that I already processed as if they were MiFish into another folder. This includes producing the sequence taxonomy file after merging and assigning taxonomy.



### Trim 5' ends of reads
All R1 should begin with the forward primer: GTCGGTAAAACTCGTGCCAGC (21 bases).
All R2 should begin with the reverse primer: CATAGTGGGGTATCTAATCCCAGTTTG (27 bases).

Trim these with the following commands:

```
for K in 24 64 81 146; do
  qiime cutadapt trim-paired \
    --i-demultiplexed-sequences trimd_Plate$K.qza \
    --p-front-f GTCGGTAAAACTCGTGCCAGC \
    --p-front-r CATAGTGGGGTATCTAATCCCAGTTTG \
    --o-trimmed-sequences trimd2_Plate$K.qza \
    --verbose > cutadapt_out2_Plate$K.txt
done
```

To see how much data passed the filter for each sample:
```
for K in 24 64 81 146; do
  grep "Total written (filtered):" cutadapt_out2_Plate$K.txt 
done
```
88% consistently for first set of samples, but then 100% for plate 146 again.

** Redo 146 for MiniFish primers**

```
for K in 146; do
  qiime cutadapt trim-paired \
    --i-demultiplexed-sequences trimd_Plate$K.qza \
    --p-front-f GTTATACGAGAGGCCCAAGTTG \
    --p-front-r TAAAGCCACTTTCGTGGTTG \
    --o-trimmed-sequences trimd2_MiniFish_Plate$K.qza \
    --verbose > cutadapt_MiniFish_out2_Plate$K.txt
done

for K in 146; do
  grep "Total written (filtered):" cutadapt_MiniFish_out2_Plate$K.txt 
done
```
This is 92% now.

Make new visualisations to see how long the sequences are now.
```
for K in 146; do
  qiime demux summarize \
    --i-data trimd2_plate$K.qza \
    --o-visualization trimd2_MiniFish_Plate$K.qzv
done
```

## 4. Denoise with dada2

Prior to denoising, sequences are trimmed to a length of 133 and 138 bp for forward and reverse reads, respectively, and merged requiring a minimum overlap of 50 base pairs to prevent spurious overlaps. We found these settings maximized the number of paired-end reads that could be merged without losing sequences due to low-quality bases being retained at the ends of reads while also preventing incorrect merging of paired-end reads with shorter overlaps. The predicted overlap for the MiFish amplicon with 250 bp paired-end sequencing and our trimming parameters is roughly 80-90 bp in length, so setting a minimum of 50 bp should be more than adequate (default is 12).

```
for K in 24 64 81 146; do
  qiime dada2 denoise-paired \
    --i-demultiplexed-seqs trimd2_Plate$K.qza \
    --p-trunc-len-f 133 \
    --p-trunc-len-r 138 \
    --p-trim-left-f 0 \
    --p-trim-left-r 0 \
    --p-min-overlap 50 \
    --p-n-threads 8 \
    --o-representative-sequences rep-seqs_Plate$K \
    --o-table table_Plate$K \
    --o-denoising-stats denoise_Plate$K
done
```

Create visualizations for the denoising stats, rep-seqs, and table. 
```
for K in 24 64 81 146; do  
  qiime metadata tabulate\
    --m-input-file denoise_Plate$K.qza\
    --o-visualization denoise_Plate$K.qzv
done

for K in 24 64 81 146; do  
  qiime feature-table tabulate-seqs \
    --i-data rep-seqs_Plate$K.qza \
    --o-visualization rep-seqs_Plate$K.qzv
done

for K in 24 64 81 146; do 
  qiime feature-table summarize \
    --i-table table_Plate$K.qza \
    --m-sample-metadata-file metadata.txt \
    --o-visualization table_Plate$K.qzv
done
```

Denoise plate 146 separately using settings for MiniFish. Trimming reads to 91bp since that is the average length of the MiniFish amplicon.
```
for K in 146; do
  qiime dada2 denoise-paired \
    --i-demultiplexed-seqs trimd2_MiniFish_Plate$K.qza \
    --p-trunc-len-f 91 \
    --p-trunc-len-r 91 \
    --p-trim-left-f 0 \
    --p-trim-left-r 0 \
    --p-min-overlap 50 \
    --p-n-threads 8 \
    --o-representative-sequences rep-seqs_MiniFish_Plate$K \
    --o-table table_MiniFish_Plate$K \
    --o-denoising-stats denoise_MiniFish_Plate$K
done

for K in 146; do  
  qiime metadata tabulate\
    --m-input-file denoise_MiniFish_Plate$K.qza\
    --o-visualization denoise_MiniFish_Plate$K.qzv
done

for K in 146; do  
  qiime feature-table tabulate-seqs \
    --i-data rep-seqs_MiniFish_Plate$K.qza \
    --o-visualization rep-seqs_MiniFish_Plate$K.qzv
done

for K in 146; do 
  qiime feature-table summarize \
    --i-table table_MiniFish_Plate$K.qza \
    --m-sample-metadata-file metadata.txt \
    --o-visualization table_MiniFish_Plate$K.qzv
done

```

## 5. Merge across plates

Not sure about merging MiFish and MiniFish amplicons, but let's try
```
qiime feature-table merge \
  --i-tables table_Plate24.qza \
  --i-tables table_Plate64.qza \
  --i-tables table_Plate81.qza \
  --i-tables table_MiniFish_Plate146.qza \
  --p-overlap-method sum \
  --o-merged-table merged-table.qza

qiime feature-table summarize \
    --i-table merged-table.qza \
    --m-sample-metadata-file metadata.txt \
    --o-visualization merged-table
    
qiime feature-table merge-seqs \
  --i-data rep-seqs_Plate24.qza \
  --i-data rep-seqs_Plate64.qza \
  --i-data rep-seqs_Plate81.qza \
  --i-data rep-seqs_MiniFish_Plate146.qza \
  --o-merged-data merged_rep-seqs.qza

qiime feature-table tabulate-seqs \
  --i-data merged_rep-seqs.qza \
  --o-visualization merged_rep-seqs.qzv
```

## 6. Assign Taxonomy

### Make a custom reference database (not run)

Below is the code I used to create a custom reference database using the RESCRIPt plugin. This database contains all fish 12S and mitochondrial sequences available on GenBank at the time of curation (June 2025), in addition to birds worked on in our lab as possible sources of contamination and/or amplification of host DNA. Note that because I made this database for a different project around the same time, I did not include Suliformes in my search, so gannet DNA ends up being either unassigned or matched to other species with low confidence.

Download sequences from GenBank, clean-up (using default settings), and dereplicate:
```
qiime rescript get-ncbi-data \
--p-query '(12s[Title] OR \
            12S[Title] OR \
            "small subunit ribosomal RNA gene"[Title] OR \
            "mitochondrion, complete genome"[Title] AND \
            ("Chondrichthyes"[Organism] OR \
            "Dipnomorpha"[Organism] OR \
            "Actinopterygii"[Organism] OR \
            "Myxini"[Organism] OR \
            "Hyperoartia"[Organism] OR \
            "Coelacanthimorpha"[Organism] OR \
            fish[All Fields] OR \
            "Sphenisciformes"[Organism] OR \
            "Charadriiformes"[Organism] OR \
            "Procellariiformes"[Organism] OR \
            "Anseriformes"[Organism]))' \
  --p-n-jobs 5 \
  --o-sequences ncbi-refseqs-unfiltered.qza \
  --o-taxonomy ncbi-taxonomy-unfiltered.qza  

qiime rescript cull-seqs \
  --i-sequences ncbi-refseqs-unfiltered.qza \
  --p-num-degenerates 5 \
  --p-homopolymer-length 8 \
  --p-n-jobs 4 \
  --o-clean-sequences ncbi-refseqs-culled
  
qiime rescript dereplicate \
  --i-sequences ncbi-refseqs-culled.qza \
  --i-taxa ncbi-taxonomy-unfiltered.qza \
  --p-mode uniq \
  --p-threads 4 \
  --o-dereplicated-sequences ncbi-refseqs-culled-derep \
  --o-dereplicated-taxa ncbi-taxonomy-culled-derep
```

Export database out of qiime:
```
mkdir editing_database

qiime tools export \
  --input-path ncbi-taxonomy-culled-derep.qza \
  --output-path editing_database
  
qiime tools export \
  --input-path ncbi-refseqs-culled-derep.qza \
  --output-path editing_database
  
```
This puts the sequences into `dna-sequences.fasta` and the taxonomy strings into `taxonomy.tsv`. I added a single human mitochondrial sequence that I downloaded from GenBank by pasting it into the fasta file and adding the corresponding taxonomy string to the taxonomy file. I did this by hand as I only need a single human sequence in my reference database to identify possible human contamination.

Import back in to QIIME2:
```
qiime tools import \
  --input-path editing_database/taxonomy.tsv \
  --output-path editing_database/ncbi-taxonomy-withHuman \
  --type 'FeatureData[Taxonomy]'
  
qiime tools import \
  --input-path editing_database/dna-sequences.fasta \
  --output-path editing_database/ncbi-refseqs-withHuman \
  --type 'FeatureData[Sequence]'  
```

### Assign taxonomy using custom python script

The custom script uses an iterative BLAST method to take each representative sequence from our samples and blast it 80 times against the custom database, increasing the percent identity incrementally from 70 – 100 %, thus circumventing the limitation of the BLAST method, which keeps only the first hit that meets the search criteria, rather than the best hit. 

The script is available [here](https://github.com/GemmaClucas/UK-Gannets/blob/main/MiFish_2024/mktaxa_singlethreaded.py).

Usage:
```
./mktaxa_singlethreaded.py \
  ncbi-refseqs-withHuman.qza \
  ncbi-taxonomy-withHuman.qza \
  merged_rep-seqs.qza

qiime metadata tabulate \
  --m-input-file superblast_taxonomy.qza \
  --o-visualization superblast_taxonomy
```

Output sequence taxonomy file with rep-seqs and their assigned taxa.
```
qiime metadata tabulate \
  --m-input-file merged_rep-seqs.qza \
  --m-input-file superblast_taxonomy.qza \
  --o-visualization sequence_taxonomy.qzv
```

Make a barplot to eye-ball what's in our samples:
```
qiime taxa barplot \
  --i-table merged-table.qza \
  --i-taxonomy superblast_taxonomy.qza \
  --m-metadata-file metadata.txt \
  --o-visualization barplot_before_filtering.qzv

```
Plate 146 looks very different to the other samples as fish are consistently being assigned to different taxa.

MiniFish primers have missed Atlantic herring in the mock community, but there is no herring in the samples from any of the other years, so this is not a problem. Are they missing other things though?

The MiniFish primers have also not amplified any guillemot DNA compared to the MiFish primers, but the amount amplified by MiFish primers is pretty minimal in most of the samples.


## 7. Remove non-prey reads

Filter out any sequences from the birds, mammals (human), and unnassigned sequences.

```
qiime taxa filter-table \
  --i-table merged-table.qza \
  --i-taxonomy superblast_taxonomy.qza \
  --p-exclude Unassigned,Aves,Mammalia \
  --o-filtered-table merged_table_noBirdsMammalsUnassigned.qza
  
qiime feature-table summarize \
    --i-table merged_table_noBirdsMammalsUnassigned.qza \
    --m-sample-metadata-file metadata.txt \
    --o-visualization merged_table_noBirdsMammalsUnassigned

qiime taxa barplot \
  --i-table merged_table_noBirdsMammalsUnassigned.qza \
  --i-taxonomy superblast_taxonomy.qza \
  --m-metadata-file metadata.txt \
  --o-visualization barplot_noBirdsMammalsUnassigned.qzv
```
Only one extraction blank and four field blanks left.

## 8. Read depth comparison

```{r}
source("/Users/gc547/Dropbox/GitHub_copied/Fecal_metabarcoding/Scripts/read_depth_analysis.R")

results <- analyze_read_depths(
  feature_table_path = "MiFish/merged_table_noBirdsMammalsUnassigned.qza",
  taxonomy_path = "MiFish/superblast_taxonomy.qza",
  metadata_path = "MiFish/metadata.txt",
  output_folder = "MiFish/", # 
  save_output = TRUE
)
```
Good news, the extraction-, field-, and PCR-banks all have barely any reads in them.

## 9. Calculate alpha rarefaction curves 

First collapsing at the species level (level 7) so that we can look at species richness within the samples, rather than ASV richness, which is the default.
```
qiime taxa collapse \
  --i-table merged_table_noBirdsMammalsUnassigned.qza \
  --i-taxonomy superblast_taxonomy.qza \
  --p-level 7 \
  --o-collapsed-table merged_table_noBirdsMammalsUnassigned_collapsed.qza

qiime diversity alpha-rarefaction \
  --i-table merged_table_noBirdsMammalsUnassigned_collapsed.qza \
  --m-metadata-file metadata.txt \
  --p-min-depth 100 \
  --p-max-depth 5000 \
  --o-visualization alpha-rarefaction-100-5000
```
Species richness reaches a plateau somewhere between 100 and 500 reads. Zoom in on that region.
```
qiime diversity alpha-rarefaction \
  --i-table merged_table_noBirdsMammalsUnassigned_collapsed.qza \
  --m-metadata-file metadata.txt \
  --p-min-depth 50 \
  --p-max-depth 600 \
  --o-visualization alpha-rarefaction-50-600
```
Species richness plateaus in the samples at 150 reads, so just drop samples with fewer than 150. This drops the final extraction blank, two of the field blanks, and two samples. Two field blanks remain with 312 and 1,102 reads.

## 10. Filtering

### Drop mock community, blanks, and samples with fewer than 150 reads

```
qiime feature-table filter-samples \
  --i-table merged_table_noBirdsMammalsUnassigned.qza \
  --p-min-frequency 150 \
  --m-metadata-file metadata.txt \
  --p-where "Type='SAMPLE'" \
  --o-filtered-table merged_table_noBirdsMammalsUnassigned_minfreq150
  
qiime taxa barplot \
  --i-table merged_table_noBirdsMammalsUnassigned_minfreq150.qza  \
  --m-metadata-file metadata.txt \
  --i-taxonomy superblast_taxonomy.qza \
  --o-visualization barplot_noBirdsMammalsUnassigned_minfreq150
```

### 1% Abundance filtering

Note that I turned this into a script when I was working on Sabiya Sheikh's 2025 gulls and shags from the UK, but I deleted it from that repo for consistency with how I had filtered her 2024 samples. But this script will now set to zero the abundance of any taxon in a sample that has an RRA less than 1% in that sample. It will also remove taxa that never reach an RRA of more than 1% in any sample.

```{r}
# Set arguments
args <- list(
  feature_table = "MiFish/merged_table_noBirdsMammalsUnassigned_minfreq150.qza",
  taxonomy = "MiFish/superblast_taxonomy.qza",
  metadata = "MiFish/metadata.txt",
  output_dir = "MiFish/",
  abundance_threshold = 0.01
)

# Source the script
source("/Users/gc547/Dropbox/GitHub_copied/Fecal_metabarcoding/Scripts/Abundance_filtering.R")

# Run
process_data(
  feature_table_path = args$feature_table,
  taxonomy_path = args$taxonomy,
  metadata_path = args$metadata,
  output_dir = args$output_dir,
  abundance_threshold = args$abundance_threshold
)
```
Import back into Qiime

```
# Import feature table
qiime tools import \
  --input-path filtered_feature_table.biom \
  --type 'FeatureTable[Frequency]' \
  --input-format BIOMV100Format \
  --output-path filtered_table_minfreq150_minabund1.qza

# Import taxonomy
qiime tools import \
  --input-path filtered_taxonomy.tsv \
  --type 'FeatureData[Taxonomy]' \
  --input-format TSVTaxonomyFormat \
  --output-path filtered_taxonomy_minfreq150_minabund1.qza
```
Make a barplot with the filtered data
```
qiime taxa barplot \
  --i-table filtered_table_minfreq150_minabund1.qza \
  --i-taxonomy filtered_taxonomy_minfreq150_minabund1.qza \
  --m-metadata-file metadata.txt \
  --o-visualization barplot_minfreq150_minabund1.qzv
```


## 11. Taxonomy edits 

Remake the sequence taxonomy file now that we have filtered out low abundance taxa.

First, filter down the rep-seqs to just those that are found in our remaining samples after abundance filtering.
```
qiime feature-table filter-seqs \
  --i-data merged_rep-seqs.qza \
  --i-table filtered_table_minfreq150_minabund1.qza \
  --o-filtered-data filtered_rep-seqs_minfreq150_minabund1.qza
```

Next make a table with sequences and taxonomy strings:
```
qiime metadata tabulate \
  --m-input-file filtered_rep-seqs_minfreq150_minabund1.qza \
  --m-input-file filtered_taxonomy_minfreq150_minabund1.qza \
  --o-visualization sequence_taxonomy_minfreq150_minabund1.qzv
```
### I THINK THIS IS RIGHT BUT CHECK WITH BIGELOW BEFORE RUNNING

Make edits after reading the taxonomy artifact into R.
```{r}
tax <- read_qza("MiFish/superblast_taxonomy.qza")

tax$data$Taxon <- tax$data$Taxon %>% 
  str_replace_all("g__Gadus;s__chalcogrammus", "g__Gadus;s__morhua") %>%       # Atlantic cod
  str_replace_all("g__Anarhichas;s__lupus", "g__Anarhichas;s__") %>%       # Wolffish sp.
  str_replace_all("g__Microcottus;s__sellaris", "g__Myoxocephalus;s__") %>%       # Sculpins sp.
  str_replace_all("g__Myoxocephalus;s__aenaeus", "g__Myoxocephalus;s__") %>%       # Sculpins sp.
  str_replace_all("g__Myoxocephalus;s__octodecemspinosus", "g__Myoxocephalus;s__") %>%       # Sculpins sp.
  str_replace_all("g__Cryptacanthodes;s__bergi", "g__;s__") %>%       # Cryptacanthodidae (wrymouths)
  str_replace_all("g__Liparis;s__miostomus", "g__Liparis;s__") %>%       # 
  str_replace_all("g__Pholis;s__crassispina", "g__Pholis;s__gunnellus") %>%       # Rock gunnell
  str_replace_all("g__Sebastes;s__mentella", "g__Sebastes;s__") %>%       # Redfish
  str_replace_all("g__Opisthocentrus;s__ocellatus", "g__Ulvaria;s__subbifurcata") %>%       # Radiated shanny 
  str_replace_all("g__Ammodytes;s__dubius", "g__Ammodytes;s__")       # Sandlance

  
  
```


Write-out the edited taxonomy file:
```{r}
if(knitr::opts_chunk$get("save_output")) {
  write.table(tax$data,
              quote = FALSE,
              row.names = FALSE,
              file = "MiFish/superblast_taxonomy_edited.tsv",
              sep = "\t")
}
```

Remove the period and reload into qiime (run in terminal). Remake barplots with updated taxonomy.
```
sed -i.bak 's/Feature.ID/Feature ID/g' superblast_taxonomy_edited.tsv

conda activate qiime2-amplicon-2024.10 

qiime tools import \
  --input-path superblast_taxonomy_edited.tsv \
  --output-path superblast_taxonomy_edited.qza \
  --type 'FeatureData[Taxonomy]'
  
qiime metadata tabulate \
  --m-input-file superblast_taxonomy_edited.qza \
  --o-visualization superblast_taxonomy_edited
  
qiime taxa barplot \
  --i-table filtered_table_minfreq150_minabund1.qza \
  --i-taxonomy superblast_taxonomy_edited.qza \
  --m-metadata-file metadata.txt \
  --o-visualization barplot_noBirdsMammalsUnassigned_minfreq150_minabund1_taxedit.qzv
```

There are some questionable similarities between consecutive samples in the 2022 batch. I think I might need to remove the top half of the samples on the plate as they all look too similar.

This will mean starting again with plate 64 after removing those samples.


### Create RRA table


```{r}
library(qiime2R)
library(dplyr)
library(tibble)

# Convert your .txt to .tsv
metadata <- read.table("MiFish/metadata.txt", sep = "\t", header = TRUE, comment.char = "#")
write.table(metadata, "MiFish/metadata.tsv", sep = "\t", row.names = FALSE, quote = FALSE)


# Version that uses the most specific taxonomy level available for each taxon
get_relative_abundance_table <- function(table_qza, metadata_tsv, taxonomy_qza, 
                                       output_file = "relative_abundance_table.csv") {
  
  feature_table <- read_qza(table_qza)$data
  taxonomy <- read_qza(taxonomy_qza)$data
  
  # Convert to relative abundance
  relative_abundance <- sweep(feature_table, 2, colSums(feature_table), FUN = "/")
  
  # Parse taxonomy
  tax_split <- taxonomy %>%
    separate(Taxon, 
            c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
            sep = ";", fill = "right", remove = FALSE) %>%
    mutate(across(Kingdom:Species, ~gsub("^[a-z]__", "", .x))) %>%
    mutate(across(Kingdom:Species, ~trimws(.x))) %>%
    # Replace empty strings and "unassigned" with NA
    mutate(across(Kingdom:Species, ~ifelse(.x == "" | .x == "unassigned", NA, .x)))
  
  # Combine relative abundance with taxonomy
  rra_with_tax <- relative_abundance %>%
    as.data.frame() %>%
    rownames_to_column("FeatureID") %>%
    left_join(tax_split, by = c("FeatureID" = "Feature.ID"))
  
  # Get sample columns (numeric columns)
  sample_cols <- rra_with_tax %>%
    select(where(is.numeric), -Consensus, -Confidence) %>%
    colnames()
  
  # Create the most specific taxonomy identifier for each row
  rra_with_tax <- rra_with_tax %>%
    rowwise() %>%
    mutate(
      # Find the most specific (last non-NA) taxonomic level
      MostSpecificTaxon = case_when(
        !is.na(Species) ~ paste(Kingdom, Phylum, Class, Order, Family, Genus, Species, sep = "; "),
        !is.na(Genus) ~ paste(Kingdom, Phylum, Class, Order, Family, Genus, sep = "; "),
        !is.na(Family) ~ paste(Kingdom, Phylum, Class, Order, Family, sep = "; "),
        !is.na(Order) ~ paste(Kingdom, Phylum, Class, Order, sep = "; "),
        !is.na(Class) ~ paste(Kingdom, Phylum, Class, sep = "; "),
        !is.na(Phylum) ~ paste(Kingdom, Phylum, sep = "; "),
        !is.na(Kingdom) ~ Kingdom,
        TRUE ~ "Unclassified"
      ),
      # Also create a label showing the taxonomic level
      TaxonomicLevel = case_when(
        !is.na(Species) ~ "Species",
        !is.na(Genus) ~ "Genus", 
        !is.na(Family) ~ "Family",
        !is.na(Order) ~ "Order",
        !is.na(Class) ~ "Class",
        !is.na(Phylum) ~ "Phylum",
        !is.na(Kingdom) ~ "Kingdom",
        TRUE ~ "Unclassified"
      )
    ) %>%
    ungroup()
  
  # Now aggregate by the most specific taxonomy
  aggregated_table <- rra_with_tax %>%
    group_by(MostSpecificTaxon, TaxonomicLevel, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
    summarise(across(all_of(sample_cols), sum, na.rm = TRUE), .groups = "drop") %>%
    arrange(TaxonomicLevel, MostSpecificTaxon)
  
  # Write to CSV
  write.csv(aggregated_table, output_file, row.names = FALSE)
  
  # Print summary
  cat("Summary of aggregated taxonomy:\n")
  print(table(aggregated_table$TaxonomicLevel))
  
  # Return the table
  return(aggregated_table)
}

# Usage:
aggregated_table <- get_relative_abundance_table("MiFish/filtered_table_minfreq150_minabund1.qza",
                                                 "MiFish/metadata.tsv", 
                                                 "MiFish/superblast_taxonomy_edited.qza",
                                                 output_file = "MiFish/relative_abundance_table.csv")

# Check the results
head(aggregated_table[, c("MostSpecificTaxon", "TaxonomicLevel")])
```
This is pretty long and the resulting table is not that easy to work with. It also loses the metadata associated with each sample, so I do not like this method.


### What about a simpler way using the csv I can get from the barplot?

```{r}
# Simple one-liner version
data <- read.csv("MiFish/level7_noBirdsMammalsUnassigned_minfreq150_minabund1_taxedit.csv")

# Find numeric columns (these are your count data)
numeric_cols <- sapply(data, is.numeric)

# Convert counts to relative abundance
data[, numeric_cols] <- sweep(data[, numeric_cols], 2, colSums(data[, numeric_cols]), FUN = "/")

# Save result
write.csv(data, "rra_table.csv", row.names = FALSE)
```
This is confusing the plate and year columns with the count data

### What about doing it from code I made for gannets?

I have modified this quite a lot to remove the 1% abundance filtering and to keep the sample metadata in the output.
```{r}
library(qiime2R)

# Read in the QIIME 2 artifacts and metadata
feature_table <- read_qza("MiFish/filtered_table_minfreq150_minabund1.qza")
taxonomy_data <- read_qza("MiFish/superblast_taxonomy_edited.qza")
metadata <- read_q2metadata("MiFish/metadata.txt")

# Convert feature table to data frame
feature_table_df <- as.data.frame(feature_table$data) %>%
  rownames_to_column("FeatureID")

# Clean up taxonomy data
taxonomy_clean <- taxonomy_data$data %>%
  parse_taxonomy() %>%  # This splits the taxonomy string into columns
  rownames_to_column("FeatureID")

# Reshape feature table to long format
feature_long <- feature_table_df %>%
  gather(key = "SampleID", value = "Abundance", -FeatureID)

# Merge all data together
complete_data <- feature_long %>%
  left_join(taxonomy_clean, by = "FeatureID") %>%
  left_join(metadata, by = "SampleID")

# Modify the species_level_data creation to handle different taxonomic levels
species_level_data <- complete_data %>%
  mutate(TaxonName = case_when(
    !is.na(Species.x) ~ paste(Genus, Species.x, sep = " "),
    !is.na(Genus) ~ paste(Genus, "(genus level)", sep = " "),
    !is.na(Family) ~ paste(Family, "(family level)", sep = " "),
    !is.na(Order) ~ paste(Order, "(order level)", sep = " "),
    !is.na(Class) ~ paste(Class, "(class level)", sep = " "),
    !is.na(Phylum) ~ paste(Phylum, "(phylum level)", sep = " "),
    TRUE ~ "Unassigned"
  )) %>%
  group_by(SampleID, TaxonName, Type, Year, Age, Colony, Plate, Species.y) %>%
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()

# Calculate relative abundance
modified_species_data <- species_level_data %>%
  group_by(SampleID) %>%
  # Calculate percentages
  mutate(TotalSampleReads = sum(Abundance),
         RelativeAbundance = (Abundance / TotalSampleReads) * 100) %>%
  ungroup()


# Pivot to wide format with samples as rows, taxa as columns this way instead
pivot_samples_wide <- function(data) {
  
  # Pivot wider: SampleID becomes rows, TaxonName becomes columns
  abundance_wide <- data %>%
    pivot_wider(id_cols = c(SampleID, Type, Year, Age, Colony, Plate, Species.y),
                names_from = TaxonName, 
                values_from = RelativeAbundance,
                values_fill = 0)
  
  return(abundance_wide)
}

# Apply the function
wide_relabund <- pivot_samples_wide(modified_species_data)

if(knitr::opts_chunk$get("save_output")) {
  write.csv(wide_relabund,
            "MiFish/relative_abundance_table.csv",
            row.names = FALSE)
}

```
This works and dovetails nicely into the next section for plotting. But I still lose the sample metadata when I pivot the data to wide format.


## 10. Quick plot
```{r}
# To keep top N taxa and group others
top_taxa_RRA <- modified_species_data %>%
  group_by(TaxonName) %>%
  summarise(total_RRA = sum(RelativeAbundance)) %>%
  arrange(desc(total_RRA)) %>%
  slice_head(n =10) %>%  # Keeping all 13 for now
  pull(TaxonName)

# stacked bar for all samples, no "Other" category as only 12 species
ggplot(modified_species_data, aes(x = SampleID, y = RelativeAbundance, fill = TaxonName)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Sample", 
       y = "Relative Abundance (%)",
       fill = "Taxon") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set3")

#ggsave("MiFish_2024/relative_abundance_per_sample.pdf", width = 12, height = 8)
```

## 11. RRA summary
```{r}
modified_species_data %>% group_by(TaxonName, Year) %>% 
  summarise(RRA_percent = mean(RelativeAbundance)) %>% 
  arrange(TaxonName, Year, -RRA_percent) %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  kable()

```
